---
title: "Group A — Yogurt Dataset: Data Dictionary"
author: "Workshop Participant"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

## Overview

This document describes every CSV file in the `data/group_a_yogurt/` folder —
what a single row represents, and what each column means.

We'll use **R** with the **tidyverse** (a collection of friendly R packages for
data science) to read and peek at each file.

```{r setup}
# Load the tidyverse — this gives us read_csv(), glimpse(), and many other tools
library(tidyverse)
```

---

## File 1: `00_sample_ids_yogurt.csv`

**What a row represents:** one biological sample collected from a study
participant at a specific time point. Each participant provided two samples:
one at *baseline* (before antibiotics) and one *after antibiotics*.

```{r file1}
sample_ids <- read_csv("../data/group_a_yogurt/00_sample_ids_yogurt.csv")

# Show the first few rows and the structure
#head(sample_ids)
#glimpse(sample_ids)
```

### Columns

| Column | Type | Meaning |
|--------|------|---------|
| `pid` | character | **Participant ID** — a unique code for each person in the study (e.g. `pid_25`). The same `pid` appears twice (baseline + after antibiotic). |
| `time_point` | character | **When the sample was taken** — either `"baseline"` (before the antibiotic course) or `"after_antibiotic"` (after the antibiotic course). |
| `arm` | character | **Study group** — which diet the participant was assigned to: `"yogurt"` (ate yogurt daily) or `"unchanged_diet"` (control group, no yogurt). |
| `sample_id` | character | **Unique sample identifier** — a code like `UNCH001` (unchanged diet) or `YOG002` (yogurt group). This is the key used to connect to the other tables. |

---

## File 2: `01_participant_metadata_yogurt.csv`

**What a row represents:** one study participant, with their background
information (demographics and health-related data).

```{r file2}
participant_metadata <- read_csv("../data/group_a_yogurt/01_participant_metadata_yogurt.csv")

head(participant_metadata)
glimpse(participant_metadata)
```

### Columns

| Column | Type | Meaning |
|--------|------|---------|
| `pid` | character | **Participant ID** — matches the `pid` in the sample IDs file. |
| `arm` | character | **Study group** — `"yogurt"` or `"unchanged_diet"`. |
| `days_since_last_sex` | double | **Days since last sexual activity** — a self-reported number. Larger numbers mean longer since last sex. |
| `birth_control` | character | **Hormonal birth control use** — either `"no hormonal birth control"` or `"Depoprovera"` (a type of injectable hormonal contraception). |
| `age` | double | **Age** of the participant in years. |
| `education` | character | **Highest education level** — categories like `"less than grade 9"`, `"grade 10-12, not matriculated"`, `"grade 10-12, matriculated"`, or `"post-secondary"`. |
| `sex` | character | **Sex of participant** — all participants are `"female"`. |

---

## File 3: `02_qpcr_results_yogurt.csv`

**What a row represents:** the bacterial DNA measurements (via qPCR — a lab
technique that counts how many copies of a gene are present) for one sample.
It tells us how much total bacteria and two specific *Lactobacillus* species
are present.

```{r file3}
qpcr <- read_csv("../data/group_a_yogurt/02_qpcr_results_yogurt.csv")

head(qpcr)
glimpse(qpcr)
```

### Columns

| Column | Type | Meaning |
|--------|------|---------|
| `sample_id` | character | **Sample ID** — links to the `sample_id` in `00_sample_ids_yogurt.csv`. |
| `qpcr_bacteria` | double | **Total bacteria** — the number of 16S rRNA gene copies detected (a standard way to measure all bacteria). Larger number = more bacteria overall. The 16S rRNA gene is a gene found in all bacteria, so it's used as a proxy for total bacterial load. |
| `qpcr_crispatus` | double | ***Lactobacillus crispatus* count** — how many copies of a gene specific to *L. crispatus* were found. This species is generally considered beneficial for vaginal health. |
| `qpcr_iners` | double | ***Lactobacillus iners* count** — how many copies of a gene specific to *L. iners* were found. *L. iners* is another common vaginal *Lactobacillus*, but it's less protective than *L. crispatus*. |

---

## File 4: `03_luminex_results_yogurt.csv`

**What a row represents:** the concentration of **one cytokine** (a type of
immune-signalling protein) measured in one sample. Because there are 10
cytokines per sample, each sample appears **10 times** in this table (once per
cytokine). This is called **long format** data.

```{r file4}
luminex <- read_csv("../data/group_a_yogurt/03_luminex_results_yogurt.csv")

head(luminex)
glimpse(luminex)
```

### Columns

| Column | Type | Meaning |
|--------|------|---------|
| `sample_id` | character | **Sample ID** — links to the `sample_id` in `00_sample_ids_yogurt.csv`. |
| `cytokine` | character | **Which cytokine was measured** — one of 10 immune-signalling proteins: `IL-1a`, `IL-1b`, `IL-6`, `IL-8`, `IL-10`, `TNFa`, `IFN-Y`, `IP-10`, `MIG`, `MIP-3a`. Each plays a different role in inflammation and immune response. |
| `conc` | double | **Concentration** — how much of that cytokine was detected in the sample (in pg/mL). Higher values usually mean more inflammation. |
| `limits` | character | **Whether the concentration was measurable** — `"within limits"` means the value is reliable. `"out of range"` means it was too low (below detection limit) or too high (above the standard curve) to measure accurately. |

---

## Putting it all together

The four files are designed to be **joined** using `sample_id` (files 1, 3, 4)
and `pid` (files 1, 2). Here's a quick look at how they connect:

```{r join-preview}
# Join sample info with qPCR results
sample_ids %>%
  inner_join(qpcr, by = "sample_id") %>%
  head(10)
```

```{r join-all}
# A full join: samples + participant metadata + qPCR + a couple of cytokines
sample_ids %>%
  inner_join(participant_metadata, by = "pid") %>%
  inner_join(qpcr, by = "sample_id") %>%
  inner_join(
    luminex %>% filter(cytokine %in% c("IL-8", "IP-10")), #combine and concentrate
    by = "sample_id"
  ) %>%
  head(10)
```

```{r}
 # Load the data first                                                            
   sample_ids <- read_csv("data/group_a_yogurt/00_sample_ids_yogurt.csv")        
   qpcr <- read_csv("data/group_a_yogurt/02_qpcr_results_yogurt.csv")            
   luminex <- read_csv("data/group_a_yogurt/03_luminex_results_yogurt.csv")      
                                                                                    getwd()
   # Now verify                                                                     
   sample_ids %>%                                                                   
     summarise(across(everything(), ~ sum(is.na(.)))) 
```

```{r}                                                                             
   file.edit("group_project/data_dictionary.qmd")                                   
 ``` 




