---
title: "R Workshop — What I've Learned So Far"
author: "Me!"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

## 1. Reading data with `read_csv()`

`read_csv()` reads a `.csv` file into R as a **data frame** (like a spreadsheet):

```{r}
library(tidyverse)

sample_ids <- read_csv("../data/group_a_yogurt/00_sample_ids_yogurt.csv")
```

---

## 2. The pipe `%>%` (pronounced "and then")

The pipe takes whatever is on the **left** and feeds it into the function on the **right** as the first argument. You read it like steps in a recipe:

```{r}
# Without pipe — inside-out, hard to read
head(arrange(filter(sample_ids, arm == "yogurt"), pid))

# With pipe — left to right, like a recipe
sample_ids %>%
  filter(arm == "yogurt") %>%
  arrange(pid) %>%
  head()
```

There's also the native pipe `|>` (built into base R), but `%>%` (from the magrittr package) is what the tidyverse uses. They do the same job.

---

## 3. `c()` — combine values

`c()` makes a collection of values:

```{r}
# A vector of numbers
c(1, 5, 10, 3)

# A vector of text
c("yogurt", "unchanged_diet")

# Using it in filter
sample_ids %>%
  filter(pid %in% c("pid_01", "pid_02", "pid_03"))
```

---

## 4. `head()` and `tail()` — peek at data

- `head()` shows the **first** 6 rows (or however many you tell it)
- `tail()` shows the **last** rows

```{r}
head(sample_ids)       # first 6 rows
head(sample_ids, 3)    # first 3 rows
```

---

## 5. `glimpse()` — see the structure

`glimpse()` shows each column, its data type, and a preview of values:

```{r}
glimpse(sample_ids)
```

---

## 6. Quarto code chunks

In a `.qmd` file, you write R code inside chunks like this:

<code>
```{r}
# Your code here
```
</code>

And the setting `code-fold: true` in the YAML header adds a **"Code" button** at the top of the HTML page — click it to show or hide all the code.

---

## 7. Joining tables

`inner_join()` combines two tables using a **key column**:

```{r}
# Join sample info with qPCR results
sample_ids %>%
  inner_join(read_csv("../data/group_a_yogurt/02_qpcr_results_yogurt.csv"), by = "sample_id") %>%
  head()
```

An inner join keeps only rows that exist in **both** tables.

---

## 8. Verifying data

Some ways to check your data makes sense:

**Check for missing values:**
```{r}
sample_ids %>%
  summarise(across(everything(), ~ sum(is.na(.))))
```

**Summary statistics (min, max, average):**
```{r}
read_csv("../data/group_a_yogurt/02_qpcr_results_yogurt.csv") %>%
  summary()
```

**Count how many in each group:**
```{r}
sample_ids %>%
  count(arm)
```

**Check all IDs link up between tables:**
```{r}
anti_join(
  read_csv("../data/group_a_yogurt/02_qpcr_results_yogurt.csv"),
  sample_ids,
  by = "sample_id"
)
# Should be 0 rows — if not, something doesn't match
```

---

## 9. Key vocabulary

| Term | Meaning |
|------|---------|
| **Data frame** | A table of data, like a spreadsheet |
| **Vector** | A collection of values (all the same type) |
| **Column** | A variable or field in the data |
| **Row** | One observation / data point |
| **Key** | A column used to link two tables together |
| **Pipe** | `%>%` — feeds data into the next step |
| **Tidyverse** | A collection of R packages for data science |

---

## 10. The yogurt dataset (Group A)

We have **4 CSV files**:

| File | One row = | Key column |
|------|-----------|------------|
| `00_sample_ids_yogurt.csv` | One biological sample | `sample_id` |
| `01_participant_metadata_yogurt.csv` | One participant | `pid` |
| `02_qpcr_results_yogurt.csv` | Bacterial DNA counts for one sample | `sample_id` |
| `03_luminex_results_yogurt.csv` | One cytokine measurement (10 rows per sample) | `sample_id` |

Study design: participants were split into **yogurt** and **unchanged diet** groups. Each person gave a sample at **baseline** and **after antibiotics**.