---
title: "Module 2 — Intro to R and Quarto"
format: html
editor: source
execute:
  warning: false
  message: false
---

This is your workspace for the Module 2 worksheet. The full instructions live
on the lesson page — keep it open alongside this notebook. You will build up
this notebook one part at a time: fill in the empty chunk under each heading,
press **Run Cell** to see what happens, and write a sentence of your own
between the chunks. Render with **Preview** whenever the worksheet tells you to.

Always load the packages you need in a setup chunk at the top. Run this first,
every time you open the notebook:

```{r setup}
library(tidyverse)
library(here)
```

## Part 1 — Run a chunk, write a sentence, render


```{r}
# 3 + 5 + 10 
10 * (5 + 1)
2 ** 10
64 / 2 + 2
64 / (2 + 2) 
```
the brackets helped to separate the equation 
the above code is a summary of multiplication and addition, 2**10 means 2 to the power of 10



## Part 2 — Read the data and look at it

```{r}
table_02 <- read_csv(here("data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))

table_02
#table_02 has 10 columns and 6 rows, of the variables, nugent_score looks the most interesting because I'm unsure what it means
glimpse(table_02)
```

## Part 3 — Pull out columns and summarise them

```{r}
table_02$nugent_score
mean(table_02$crp_blood, na.rm = TRUE) #na.rm = TRUE: skip missing numbers
mean(table_02$crp_blood) #the numbers stayed the same
crp_mean <- mean(table_02$crp_blood, na.rm = TRUE)
crp_mean
#crp_mean is the mean value of the column crp_blood and "na.rm = TURE" ensures that missing values are not counted
```

## Part 4 — Build something with Pi

```{r}
# your code here
```

## Capstone — make it your own

```{r}
# your code here
```

## Before you go 
