---
title: "Group A Yogurt — Results"
format: html
editor: source
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(gtsummary)
library(here)
library(ggpubr)
```

```{r}
# Load the Group A Yogurt participant data
participants <- read_csv(here::here("data/group_a_yogurt/01_participant_metadata_yogurt.csv"))
```

```{r table1}
# Build Table 1 — descriptive stats by arm (yogurt vs unchanged diet)
participants |>
  select(-pid, -sex) |>                      # drop ID and sex (all female)
  tbl_summary(
    by = arm,                                 # split table by diet group
    statistic = list(age ~ "{mean} ({sd})")   # show age as mean (SD)
  ) |>
  add_p() |>                                  # add p-values
  add_overall() |>                            # add an overall column
  bold_labels() |>                            # bold variable names
  modify_header(label ~ "**Characteristic**") # nicer column name
```

```{r boxplot}
# Boxplot: age by birth control type, faceted by diet arm
# facet_wrap(~ arm) splits into separate panels for yogurt vs unchanged diet
# stat_compare_means() adds a p-value (stars) on each panel

ggplot(participants, aes(x = birth_control, y = age)) +
  geom_boxplot(aes(fill = arm)) +
  facet_wrap(~ arm) +                        # one panel per diet group
 ggpubr::stat_compare_means(                         # add p-value stars
    method = "wilcox.test",
    label = "p.signif"
  ) +
  labs(
    title = "Age by Birth Control and Diet Group",
    x = "Birth Control",
    y = "Age (years)",
    fill = "Diet Group"
  ) +
  theme_minimal()
```

---

## Bonus: Using qPCR or Luminex data in boxplots

### 🧬 qPCR data — `02_qpcr_results_yogurt.csv`

qPCR measures **bacterial DNA** levels in each sample. It's already in **wide format**
(one row per sample, each column is a continuous number).

| Column | What it measures |
|---|---|
| `qpcr_bacteria` | **Total bacteria** — all bacteria combined |
| `qpcr_crispatus` | ***Lactobacillus crispatus*** — a "good" bacteria linked to vaginal health |
| `qpcr_iners` | ***Lactobacillus iners*** — another common Lactobacillus species |

### 🧪 Luminex data — `03_luminex_results_yogurt.csv`

Luminex measures **immune markers (cytokines)** — proteins your body makes when
it's inflamed or fighting infection. This data is in **long format** (each row =
one sample + one cytokine). To use it in a boxplot, you pick **one cytokine at a time**.

The 10 cytokines available:

| Cytokine | What it roughly indicates |
|---|---|
| **IL-1a, IL-1b, IL-6, TNFa** | Inflammation signals |
| **IL-8, IP-10, MIG, MIP-3a** | Chemical signals that attract immune cells |
| **IL-10** | Anti-inflammatory (calms things down) |
| **IFN-Y** | Antiviral / immune activation |

### 🔗 Joining the data together

To make boxplots with qPCR/Luminex data and birth control info, we need to join
three files together using `sample_id` as the link:

```
sample_ids ───▶ has arm + time_point for each sample_id
                    │
                    ▼ joins by sample_id
qPCR or Luminex ───▶ adds the bacterial/cytokine measurements
                    │
participant meta ───▶ adds birth_control (joins by pid)
```

```{r load_all_data}
# Load sample IDs (links pid to sample_id, arm, time_point)
sample_ids <- read_csv(here::here("data/group_a_yogurt/00_sample_ids_yogurt.csv"))

# Load qPCR data (bacterial measurements per sample)
qpcr <- read_csv(here::here("data/group_a_yogurt/02_qpcr_results_yogurt.csv"))

# Load Luminex data (cytokine measurements per sample)
luminex <- read_csv(here::here("data/group_a_yogurt/03_luminex_results_yogurt.csv"))

# Join everything together
# First join qPCR with sample_ids to get arm and time_point
qpcr_joined <- qpcr |>
  left_join(sample_ids, by = "sample_id") |>
  left_join(participants, by = c("pid", "arm"))  # add birth_control

# For Luminex, pick one cytokine and join the same way
luminex_joined <- luminex |>
  left_join(sample_ids, by = "sample_id") |>
  left_join(participants, by = c("pid", "arm"))
```

```{r qpcr_boxplot}
# Example 1: Boxplot of total bacteria by birth control, faceted by arm
# (y-axis shows bacterial DNA levels)
ggplot(qpcr_joined, aes(x = birth_control, y = qpcr_bacteria)) +
  geom_boxplot(aes(fill = arm)) +
  facet_wrap(~ arm) +
  ggpubr::stat_compare_means(method = "wilcox.test", label = "p.signif") +
  scale_y_log10() +                           # log scale helps see the spread
  labs(
    title = "Total Bacteria by Birth Control and Diet Group",
    x = "Birth Control",
    y = "Total Bacteria (log scale)",
    fill = "Diet Group"
  ) +
  theme_minimal()
```

```{r luminex_boxplot}
# Example 2: Boxplot of IL-8 (inflammation signal) by birth control, faceted by arm
# Filter to just one cytokine first
luminex_joined |>
  filter(cytokine == "IL-8") |>
  ggplot(aes(x = birth_control, y = conc)) +
  geom_boxplot(aes(fill = arm)) +
  facet_wrap(~ arm) +
  ggpubr::stat_compare_means(method = "wilcox.test", label = "p.signif") +
  labs(
    title = "IL-8 Levels by Birth Control and Diet Group",
    x = "Birth Control",
    y = "IL-8 Concentration",
    fill = "Diet Group"
  ) +
  theme_minimal()
```

### 🎮 Try changing these:

| In the qPCR plot, swap to... | To look at... |
|---|---|
| `y = qpcr_crispatus` | L. crispatus (beneficial bacteria) |
| `y = qpcr_iners` | L. iners (another species) |

| In the Luminex plot, change `filter(cytokine == ...)` to... | To look at... |
|---|---|
| `"IL-6"` or `"TNFa"` | Different inflammation signals |
| `"IL-10"` | Anti-inflammatory signal |
| `"IP-10"` or `"MIG"` | Immune cell recruitment signals |

---

## Regression Models

### 🤔 What is regression, in plain words?

Think of regression as asking:

> *"If I change **one thing** (like eating yogurt), how much does **another thing** (like bacteria levels) change — **while holding everything else constant**?"*

A boxplot shows you a raw comparison. But regression lets you **control for confounders** like birth control and age. So you can ask: *"Is the yogurt effect real, or is it just because the yogurt group has more Depoprovera users?"*

### 📐 The model formula

```
outcome ~ arm + birth_control + age + time_point
```

This says: "Predict the outcome using diet group, while **adjusting for** birth control, age, and time point."

The coefficient for `arm` tells us: *"How much does the yogurt group differ from the unchanged diet group, after accounting for birth control, age, and time?"*

### 🧬 Regression 1: Total Bacteria

Bacterial counts span a huge range (0 to 80 million), so we use `log()` to make them behave better.

```{r qpcr_regression}
# Linear regression: does diet group predict total bacteria,
# after controlling for birth control, age, and time point?

qpcr_model <- lm(log(qpcr_bacteria + 1) ~ arm + birth_control + age + time_point,
                  data = qpcr_joined)

# View the results
summary(qpcr_model)
```

```{r qpcr_regression_table}
# Cleaner table using gtsummary
qpcr_model |>
  tbl_regression() |>
  bold_labels()
```

### 🧪 Regression 2: IL-8 (Inflammation)

```{r luminex_regression}
# Filter to just IL-8, then run the same model
il8_data <- luminex_joined |>
  filter(cytokine == "IL-8")

il8_model <- lm(log(conc + 1) ~ arm + birth_control + age + time_point,
                data = il8_data)

summary(il8_model)
```

```{r luminex_regression_table}
il8_model |>
  tbl_regression() |>
  bold_labels()
```

### 🧪 Regression 3: Try any other cytokine

Just change the cytokine name in the `filter()` line!

```{r cytokine_of_choice}
# Try IL-6, TNFa, IP-10, IL-10, or any of the 10 cytokines
my_cytokine <- "IL-6"

cyt_data <- luminex_joined |>
  filter(cytokine == my_cytokine)

cyt_model <- lm(log(conc + 1) ~ arm + birth_control + age + time_point,
                data = cyt_data)

cyt_model |>
  tbl_regression() |>
  bold_labels() |>
  modify_caption(paste("Regression for", my_cytokine))
```

### 📖 How to read the regression output

```
Coefficients:
                       Estimate  Pr(>|t|)    
(Intercept)            5.79e+00  < 5e-16 ***  ← baseline (unchanged diet, no BC, baseline time)
armyogurt              6.40e-02    0.755      ← yogurt effect (ns)  
birth_controlDepo      1.30e+00   < 0.001 *** ← Depoprovera effect (significant!)
age                   -5.93e-02    0.034 *    ← age effect (older = slightly less bacteria)
time_pointafter...     5.28e-01    0.004 **   ← time effect (more bacteria after antibiotics)
```

| Term | What it means |
|---|---|
| `(Intercept)` | The predicted value when all other variables are 0 (the baseline) |
| `armyogurt` | **The yogurt effect** — difference between yogurt & unchanged diet, holding everything else constant. P > 0.05 = not significant |
| `birth_controlDepo` | Difference between Depoprovera & no hormonal BC. If significant, birth control matters! |
| `age` | How much bacteria changes for each extra year of age |
| `time_pointafter_antibiotic` | Difference between after-antibiotic & baseline |

### 💡 Key insight

Notice how regression **controls for birth control**. The boxplots showed raw differences, but regression asks:

> *"If I compare two women of the **same age**, with the **same birth control**, at the **same time point** — does eating yogurt make a difference?"*

This is much more powerful than just looking at boxplots!

### 🧠 Something to think about

Each participant was measured **twice** (baseline + after antibiotics). Regression treats those as independent, but they're really linked. A more advanced approach would be a **mixed model** or looking at **change from baseline**:

```{r change_from_baseline}
# Calculate change from baseline for each participant
qpcr_change <- qpcr_joined |>
  select(pid, arm, time_point, qpcr_bacteria, birth_control, age) |>
  pivot_wider(names_from = time_point, values_from = qpcr_bacteria) |>
  mutate(bacteria_change = log(after_antibiotic + 1) - log(baseline + 1))

# Now model the change (each participant appears only once)
change_model <- lm(bacteria_change ~ arm + birth_control + age, data = qpcr_change)

change_model |>
  tbl_regression() |>
  bold_labels() |>
  modify_caption("Predicting **change** in bacteria from baseline to after antibiotics")
```

This tells you: *"Does eating yogurt affect how much your bacteria **changes** from before to after antibiotics?"*
```