---
title: "Group A Yogurt — Data Summary"
subtitle: "Effects of yogurt on the vaginal microbiome during antibiotics"
author: "Workshop Group A"
format: 
  revealjs:
    theme: dark
    transition: slide
    slide-number: true
    preview-links: auto
execute:
  warning: false
  message: false
editor: source
---

## The Study Design

- **Question:** Does eating yogurt during antibiotic treatment affect the vaginal microbiome?
- **Participants:** 51 women
- **Two diet groups:**
  - 🥛 **Yogurt group** — 26 women
  - 🚫 **Unchanged diet group** — 25 women
- **Two time points:** Baseline → then after antibiotics
- **Measurements:** Demographics, bacterial DNA (qPCR), immune markers (Luminex)

---

## The Data Files

| File | What's inside |
|---|---|
| `00_sample_ids_yogurt.csv` | Links each sample to a participant, time point & diet group |
| `01_participant_metadata_yogurt.csv` | Age, birth control, education, etc. |
| `02_qpcr_results_yogurt.csv` | Bacterial DNA levels (total, L. crispatus, L. iners) |
| `03_luminex_results_yogurt.csv` | 10 immune markers (cytokines) per sample |

---

## Step 1: Table 1 — Participant Characteristics

We compared demographics between the yogurt and unchanged diet groups.

| Characteristic | Unchanged Diet (N=25) | Yogurt (N=26) | p-value |
|---|---|---|---|
| Days since last sex | 6 (3, 10) | 5 (3, 11) | 0.8 |
| **Birth control** 🚨 | | | **0.017*** |
| &nbsp;&nbsp;Depoprovera | 8 (32%) | **17 (65%)** | |
| &nbsp;&nbsp;No hormonal BC | 17 (68%) | 9 (35%) | |
| Age | 30.4 (3.3) | 31.0 (3.3) | 0.6 |
| Education | — | — | 0.8 |

---

## Table 1 — Key Finding

**Birth control is unbalanced!** 🚨

- The yogurt group has **65% Depoprovera users**
- The unchanged diet group has only **32% Depoprovera users**
- This difference is statistically significant (p = 0.017)

This means birth control could be a **confounder** — any difference we see between diet groups might actually be due to birth control differences, not yogurt.

---

## Step 2: Boxplot — Age by Birth Control

```{r boxplot_again, echo=FALSE}
library(tidyverse)
library(here)
library(ggpubr)

participants <- read_csv(here::here("data/group_a_yogurt/01_participant_metadata_yogurt.csv"))

ggplot(participants, aes(x = birth_control, y = age)) +
  geom_boxplot(aes(fill = arm)) +
  facet_wrap(~ arm) +
  ggpubr::stat_compare_means(method = "wilcox.test", label = "p.signif") +
  labs(
    title = "Age by Birth Control and Diet Group",
    x = "Birth Control",
    y = "Age (years)",
    fill = "Diet Group"
  ) +
  theme_minimal()
```

---

## Age by Birth Control — Interpretation

| Group | Depoprovera (avg) | No hormonal BC (avg) | p-value |
|---|---|---|---|
| **Unchanged diet** | 32.6 years | 29.4 years | **0.022*** |
| **Yogurt** | 32.0 years | 29.1 years | 0.061 (ns) |

**What this means:**

- In both groups, Depoprovera users are **~3 years older** on average
- This is significant in the unchanged diet group (\*)
- Since birth control is unbalanced between arms, **age differences could be driving some of our results**

---

## Step 3: Boxplot — Total Bacteria (qPCR)

```{r qpcr_boxplot_again, echo=FALSE}
sample_ids <- read_csv(here::here("data/group_a_yogurt/00_sample_ids_yogurt.csv"))
qpcr <- read_csv(here::here("data/group_a_yogurt/02_qpcr_results_yogurt.csv"))

qpcr_joined <- qpcr |>
  left_join(sample_ids, by = "sample_id") |>
  left_join(participants, by = c("pid", "arm"))

ggplot(qpcr_joined, aes(x = birth_control, y = qpcr_bacteria)) +
  geom_boxplot(aes(fill = arm)) +
  facet_wrap(~ arm) +
  ggpubr::stat_compare_means(method = "wilcox.test", label = "p.signif") +
  scale_y_log10() +
  labs(
    title = "Total Bacteria by Birth Control and Diet Group",
    x = "Birth Control",
    y = "Total Bacteria (log scale)",
    fill = "Diet Group"
  ) +
  theme_minimal()
```

---

## Total Bacteria — Interpretation

| Group | Depoprovera (median) | No hormonal BC (median) | p-value |
|---|---|---|---|
| **Unchanged diet** | 4.8 million | 1.5 million | 0.084 (ns) |
| **Yogurt** | 2.0 million | 2.9 million | 0.736 (ns) |

**Takeaways:**

- In the unchanged diet group, Depoprovera users have **3× more bacteria** — but not significant (p = 0.084)
- In the yogurt group, bacteria levels are similar regardless of birth control
- **Suggestion:** Could yogurt be "smoothing out" differences caused by birth control?

---

## Step 4: Boxplot — IL-8 (Inflammation)

```{r luminex_boxplot_again, echo=FALSE}
luminex <- read_csv(here::here("data/group_a_yogurt/03_luminex_results_yogurt.csv"))

luminex_joined <- luminex |>
  left_join(sample_ids, by = "sample_id") |>
  left_join(participants, by = c("pid", "arm"))

luminex_joined |>
  filter(cytokine == "IL-8") |>
  ggplot(aes(x = birth_control, y = conc)) +
  geom_boxplot(aes(fill = arm)) +
  facet_wrap(~ arm) +
  ggpubr::stat_compare_means(method = "wilcox.test", label = "p.signif") +
  labs(
    title = "IL-8 Levels by Birth Control and Diet Group",
    x = "Birth Control",
    y = "IL-8 Concentration",
    fill = "Diet Group"
  ) +
  theme_minimal()
```

---

## IL-8 — Interpretation

| Group | Depoprovera (median) | No hormonal BC (median) | p-value |
|---|---|---|---|
| **Unchanged diet** | 52.5 | 17.1 | 0.094 (ns) |
| **Yogurt** | 25.3 | 30.7 | 0.544 (ns) |

**What this means:**

- IL-8 is an **inflammation signal** — higher values can mean more inflammation
- In the unchanged diet group, Depoprovera users have **3× higher IL-8** (trend, not significant)
- In the yogurt group, IL-8 is similar regardless of birth control

---

## Summary of Findings

### 🎯 What we discovered

1. **Birth control is unbalanced** between diet groups — more Depoprovera users in the yogurt group (p = 0.017)

2. **Depoprovera users tend to be older** — by about 3 years on average

3. **Depoprovera users (in the unchanged diet group) show hints of:**
   - More total bacteria (ns)
   - Higher IL-8 inflammation signal (ns)

4. **In the yogurt group**, these differences seem smaller — birth control groups look more similar

---

## Key Takeaway

### 🧐 Why this matters

When comparing yogurt vs unchanged diet, we need to be careful:

> "Are we seeing an effect of **yogurt**, or are we seeing an effect of **birth control** (which happens to be unbalanced between groups)?"

### 💡 Next steps

- Compare only **Depoprovera users** across both arms
- Compare only **no-hormonal-BC users** across both arms
- Or use a statistical model that **adjusts for birth control**

---

## What We Used

**R Packages:**
| Package | Purpose |
|---|---|
| `tidyverse` | Data wrangling & ggplot2 plots |
| `gtsummary` | Beautiful Table 1 with p-values |
| `ggpubr` | Adding p-values to plots |
| `here` | File paths |

**Statistical tests:** Wilcoxon rank-sum test (non-parametric, safe for small groups)

---

# Thank you! 🥛📊

Questions? Time to explore more cytokines or bacteria!